---
title: "Data Analytics Practical 3"
author: "Eryk Gloginski"
date: "23/02/2023"
output: html_document
---

```{r create and set directory}
getwd()
dir.create("D:/Year 3/Semester 6/Data Analytics/Practicals/Practical3")
setwd("D:/Year 3/Semester 6/Data Analytics/Practicals/Practical3")
getwd()
list.files()
```
# Create and set Directory

```{r install tidyverse}
# install.packages("tidyverse")
library(tidyverse)

```
# Install and load tidyverse

```{r demo of a forward pipe}
# create a simple function for the equation: x3+3
f<- function(x)
  x^3+3

# Then we can apply f to an argument x using
x <- 5
f(x)

# The forward-pipes from magrittr allow us to rewrite this function call as
# install.packages("magrittr")
library(magrittr)
x %>% f
```
# Demo of a forward pipe where you have to install magrittr package

```{r Pipelines 2-5}
library(tidyverse)
library(magrittr)
# mtcars2 <- transform(mtcars, lp100k=235.21/mpg) subset(mtcars2, lp100k<=10)

subset(transform(mtcars, lp100k=235.21/mpg), lp100k<=10)

mtcars %>% transform(lp100k=235.21/mpg) %>% subset(lp100k<=10)

```
# Pipelines continued 2 to 5. Also transforming mtcars was not working into mtcars2 without the pipeline.

```{r Exercise 1 A & B}
library(tidyverse)
library(magrittr)

# Convert it to nested function calls. 
rnorm(1000) %>% sin() %>% max() 
# Converted Code:
max(sin(rnorm(1000)))

library(MASS) #Load package MASS, which contains the data
# Convert the R code below to pipelines.
subset(transform(mammals, ratio=brain/body), ratio==max(ratio))
# Converted Code:
mammals %>% transform(ratio=brain/body) %>% subset(ratio==max(ratio))
```
# Exercise 1. A and B, code from no pipelines to with pipelines.

```{r Pipelines 6-9}
library(tidyverse)
library(magrittr)
# Part 9
mtcars2 <- mtcars %>% transform(lp100k=235.21/mpg) %>% subset(lp100k<=10)
cor(mtcars2$disp, mtcars2$hp)

# Part 10
mtcars %>% transform(lp100k=235.21/mpg) %>% subset(lp100k<=10) %$% cor(disp, hp)

```
# Using Pipeline with additional operator.

```{r Tibbles 10-19}
library(tidyverse)
library(magrittr)
# starwars
starwars[,c("name", "starships")] #Print only these columns
starwars[10,"starships"][[1]] #prints starships flown by Obi-Wan

```
# Tibles 11 - 19.

```{r Creating Tibbles 20-27}
# Part 21
library(tidyverse)
library(magrittr)
kids <- rbind(c( 4, 15, 101),
c(11, 28, 132))
colnames(kids) <- c("age", "weight", "height")
rownames(kids) <- c("Sarah", "John")
kids <- as.data.frame(kids)
kids <- cbind(kids, gender=c("f","m"))
kids

# Part 22
kidstibble <- as_tibble(kids)
kidstibble

# Part 24
kidstibble$name <- rownames(kids) # Create a column with names
# Next we re-arrange columns so that names comes first
kidstibble <- kidstibble[,c("name", "age", "weight", "height", "gender")]
kidstibble

# Part 25
kidstibble2 <- rownames_to_column(kids)
kidstibble2 

# Part 26
kidsdf <- as.data.frame(kidstibble)
kidsdf
```
# Creating Tibles 20-27

```{r Using readr 28-31}
library(tidyverse)
library(magrittr)
read_csv("chol.csv", na=".")
```
# Using Readr 28-31

```{r Exercise 2}
library(tidyverse)
library(magrittr)
read_csv("cars.csv", na=".")
read_delim("ships.txt", na=".")

```
# Exercise 2 where we read 2 data files cars.csv and ships.txt using read_csv() and read_delim(). The na="." basically changes all . to not available.

```{r Using dplyr 32-45}
library(tidyverse)
library(magrittr)

# Part 37
load("velib.RData")

# Part 40
stationsparis <- stations %>% filter(departement=="Paris")
stationsparis

# Part 42
stations7592 <- stations %>% filter(departement=="Paris" | departement=="Hauts-de-Seine")
stations7592

# Part 43
stations7593 <- stations %>% filter(departement %in% c("Paris" , "Hauts-de-Seine"))
stations7593

# Part 44
stations7594 <- filter(stations, departement %in% c("Paris" , "Hauts-de-Seine"))
stations7594

# Part 45
stations %>% slice(5:7)
```
# Using dplyr 32-45

```{r Exercise 3}
library(tidyverse)
library(magrittr)
bikes

bikes7592 <- bikes %>% filter(available_bikes
>60 & time>15)
bikes7592
```
# Exercise 3 where I find stations that have more than 60 bikes available at 3 pm(time value 15). Had to figure out that bikes is a seperate data set.

```{r Using mutate 46-50}
library(tidyverse)
library(magrittr)

# Part 47
bikes <- bikes %>% mutate(total_stands = available_bikes+available_bike_stands)
bikes
```
# Using mutate 46-50

```{r Using arrange 51-53}
library(tidyverse)
library(magrittr)

# Part 52
bikes %>% arrange(available_bikes)

# Part 53
bikes %>% arrange(desc(available_bikes))
```
# Using Arrange 51-53

```{r Exercise 4}
library(tidyverse)
library(magrittr)
stations

stations %>% arrange(lng) %>% slice(1:3)
```
# Exercise 3 where we identify 3 bike stations that are the furthest to the West(smallest longtitude)

```{r Using group_by and summarise 54-55}
library(tidyverse)
library(magrittr)
bikes %>% group_by(name) %>% #Group by station name
summarise(avg_stands=mean(available_bike_stands)) %>% #averages
arrange(desc(avg_stands)) #Sort in descending order
```
# Using group_by and summarise 54-55

```{r Exercise 5}
library(tidyverse)
library(magrittr)

bikes
bikes %>% group_by(name) %>% summarise(n = n())
```
# Unable to finish this as I don't understand this

```{r Using join-type functions 56-60}
library(tidyverse)
library(magrittr)

bikes %>% inner_join(stations) # Merge data (using common variable: name)
```
# Using join-type functions 56-60

```{r Exercise 6}
library(tidyverse)
library(magrittr)

# Part A
load("patients_weights.Rdata")
patientweights <- weights %>% inner_join(patients)
patientweights

# Part B
patientweights %>% group_by(Gender) %>% summarise(avg_weight=mean(Weight))

```
# Exercise 6 where we merge data and get the average weight of female and male patients
